---
title: "Comparing different cluster structures"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries we are going to use.

```{r}
library(data.table)
library(tidyverse)
library(dbscan)
library(scales)
library(ggthemes)
library(ggtree)

library(coala)
activate_ms(priority = 2000)
list_simulators()

library(reticulate)
np <- reticulate::import('numpy')

pcadims <- 10

colours <- c("#ff7f00","#cab2d6","#6a3d9a","#fdbf6f","#ffff99","#1f78b4","#b15928","#33a02c","#e31a1c","#a6cee3","#fb9a99","#b2df8a")
set.seed(123)
```

Here we compare PCA, t-SNE and Mandrake on different population structures.

### No structure

We simulate datatsets with very large rates of recombination which breaks up the underlying population structure.

```{r}
model <- coala::coal_model(sample_size = 100, loci_number = 1, loci_length = 1000) +
    feat_mutation(10) +
    feat_recombination(1000) +
    sumstat_seg_sites("segsites") +
    sumstat_trees()
 
simu.stats <- simulate(model)

d <- dist(as.matrix(simu.stats$segsites[[1]])>0, method = 'binary')

njt_high_recombination <- ape::bionj(d)
plot(njt_high_recombination)

d <- as.matrix(d)
heatmap(d)

#save to numpy format for Mandrake
np$savez("./data/simulations/high_recombination.npz", 
         I=matrix(rep(seq(0, nrow(d)-1), nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         J=matrix(rep(seq(0, nrow(d)-1), each=nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         dists=d[upper.tri(d)], 
         names=paste('sample',1:nrow(d), sep='_'))
```

```
mandrake --distances ./data/simulations/high_recombination.npz --output ./data/simulations/high_recombination --maxIter 1000000
```

```{r}
mandrake_embedding_high_recombination <- fread("./data/simulations/high_recombination.embedding.txt") %>% as_tibble()

pca_high_recombination <-irlba::prcomp_irlba(d, n=pcadims)

plot(pca_high_recombination$x)

tsne_embedding_high_recombination <- Rtsne::Rtsne(X = pca_high_recombination$x, 
                  check_duplicates=FALSE,
                  pca = FALSE,
                  perplexity = 15)

plot(tsne_embedding_high_recombination$Y)

umap_embedding_high_recombination <- umap::umap(d = pca_high_recombination$x)

plot(umap_embedding_high_recombination$layout)
```


### Clines

```{r}
ngenomes <- 200

sim <- scrm::scrm(paste(c(ngenomes, ' 1 ', 
  paste(paste('-eI', 1:ngenomes*0.1), '1 '),
  ' -r 100 100 -t 1.5 -T'), collapse = ''))

d <- dist(as.matrix(sim$seg_sites[[1]])>0, method = 'binary')

njt_cline <- ape::bionj(d)
plot(njt_cline)

d <- as.matrix(d)
heatmap(d)

#save to numpy format for Mandrake
np$savez("./data/simulations/cline.npz", 
         I=matrix(rep(seq(0, nrow(d)-1), nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         J=matrix(rep(seq(0, nrow(d)-1), each=nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         dists=d[upper.tri(d)], 
         names=paste('sample',1:nrow(d), sep='_'))
```

```
mandrake --distances ./data/simulations/cline.npz --output ./data/simulations/cline --maxIter 1000000
```

```{r}
mandrake_embedding_cline <- fread("./data/simulations/cline.embedding.txt") %>% as_tibble()

plot(mandrake_embedding_cline)

pca_cline <-irlba::prcomp_irlba(d, n=pcadims)

plot(pca_cline$x)

tsne_embedding_cline <- Rtsne::Rtsne(X = pca_cline$x, 
                  check_duplicates=FALSE,
                  pca = FALSE,
                  perplexity = 15)

plot(tsne_embedding_cline$Y)

umap_embedding_cline <- umap::umap(d = pca_cline$x)

plot(umap_embedding_cline$layout)
```

### Clades

```{r}
model <- coala::coal_model(sample_size = c(50, 50, 50, 50, 50), loci_number = 1, loci_length = 1000) +
    feat_mutation(10) +
    feat_recombination(10) +
    feat_migration(rate = 1, symmetric = TRUE, time = 1) +
    feat_migration(rate = 0, symmetric = TRUE, time = 0.1) +
    sumstat_seg_sites("segsites") +
    sumstat_trees()
 
simu.stats <- simulate(model)

d <- dist(as.matrix(simu.stats$segsites[[1]])>0, method = 'binary')

njt_clade <- ape::bionj(d)
plot(njt_clade)

d <- (as.matrix(d)-mean(d))/max(d)
heatmap(d)

#save to numpy format for Mandrake
np$savez("./data/simulations/clades.npz", 
         I=matrix(rep(seq(0, nrow(d)-1), nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         J=matrix(rep(seq(0, nrow(d)-1), each=nrow(d)), nrow = nrow(d))[upper.tri(d)],
         dists=d[upper.tri(d)], 
         names=paste('sample',1:nrow(d), sep='_'))
```

```
mandrake --distances ./data/simulations/clades.npz --output ./data/simulations/clades --maxIter 1000000
```

```{r}
mandrake_embedding_clades <- fread("./data/simulations/clades.embedding.txt") %>% as_tibble()
plot(mandrake_embedding_clades)

pca_clades <-irlba::prcomp_irlba(d, n=pcadims)
plot(pca_clades$x)

tsne_embedding_clades <- Rtsne::Rtsne(X = pca_clades$x, 
                  check_duplicates=FALSE,
                  pca = FALSE,
                  perplexity = 15)
plot(tsne_embedding_clades$Y)

umap_embedding_clades <- umap::umap(d = pca_clades$x)
plot(umap_embedding_clades$layout)
```


### Variation at different resolutions

Similar to two different plasmids with different gene contents

```{r}
model <- coala::coal_model(sample_size = c(50, 50, 50, 50, 50), loci_number = 1, loci_length = 1000) +
    feat_mutation(10) +
    feat_recombination(10) +
    feat_migration(rate = 1, symmetric = TRUE, time = 1) +
    feat_migration(rate = 0, symmetric = TRUE, time = 0.1) +
    feat_growth(rate = 1000, population = 1, time = 0.05) +
    feat_growth(rate = 1000, population = 2, time = 0.05) +
    sumstat_seg_sites("segsites") +
    sumstat_trees()
 
simu.stats <- simulate(model)

d <- dist(as.matrix(simu.stats$segsites[[1]])>0, method = 'binary')

njt_clade <- ape::bionj(d)
plot(njt_clade)

d <- (as.matrix(d)-mean(d))/max(d)
heatmap(d)

#save to numpy format for Mandrake
np$savez("./data/simulations/diff_res.npz", 
         I=matrix(rep(seq(0, nrow(d)-1), nrow(d)), nrow = nrow(d))[upper.tri(d)], 
         J=matrix(rep(seq(0, nrow(d)-1), each=nrow(d)), nrow = nrow(d))[upper.tri(d)],
         dists=d[upper.tri(d)], 
         names=paste('sample',1:nrow(d), sep='_'))
```

```
mandrake --distances ./data/simulations/diff_res.npz --output ./data/simulations/diff_res --maxIter 1000000 --perplexity 50
```

```{r}
mandrake_embedding_diff_res <- fread("./data/simulations/diff_res.embedding.txt") %>% as_tibble()
plot(mandrake_embedding_diff_res)

pca_diff_res <-irlba::prcomp_irlba(d, n=pcadims)
plot(pca_diff_res$x)

tsne_embedding_diff_res <- Rtsne::Rtsne(X = pca_diff_res$x, 
                  check_duplicates=FALSE,
                  pca = FALSE,
                  perplexity = 50)
plot(tsne_embedding_diff_res$Y)

umap_embedding_diff_res <- umap::umap(d = pca_diff_res$x)
plot(umap_embedding_diff_res$layout)
```

### Summary plots

```{r}
all_embeddings <- rbind(
  rbind(
    as.matrix(mandrake_embedding_clades),
    as.matrix(pca_clades$x[,1:2]),
    as.matrix(tsne_embedding_clades$Y),
    as.matrix(umap_embedding_clades$layout)
  ) %>% as_tibble() %>% 
    add_column(sim='Five distinct clades') %>%
    add_column(method=rep(c('Mandrake','PCA','tSNE','UMAP'), each=nrow(mandrake_embedding_clades))),
  
  rbind(
    as.matrix(mandrake_embedding_high_recombination),
    as.matrix(pca_high_recombination$x[,1:2]),
    as.matrix(tsne_embedding_high_recombination$Y),
    as.matrix(umap_embedding_high_recombination$layout)
  ) %>% as_tibble() %>% 
    add_column(sim='No structure (high recombination)') %>%
    add_column(method=rep(c('Mandrake','PCA','tSNE','UMAP'), each=nrow(mandrake_embedding_high_recombination))),
  
  rbind(
    as.matrix(mandrake_embedding_cline),
    as.matrix(pca_cline$x[,1:2]),
    as.matrix(tsne_embedding_cline$Y),
    as.matrix(umap_embedding_cline$layout)
  ) %>% as_tibble() %>% 
    add_column(sim='Uniform sample times (cline)') %>%
    add_column(method=rep(c('Mandrake','PCA','tSNE','UMAP'), each=nrow(mandrake_embedding_cline))),
  
  rbind(
    as.matrix(mandrake_embedding_diff_res),
    as.matrix(pca_diff_res$x[,1:2]),
    as.matrix(tsne_embedding_diff_res$Y),
    as.matrix(umap_embedding_diff_res$layout)
  ) %>% as_tibble() %>% 
    add_column(sim='Five clades with two different growth rates') %>%
    add_column(method=rep(c('Mandrake','PCA','tSNE','UMAP'), each=nrow(mandrake_embedding_diff_res)))
)


pp <- imap(split(all_embeddings, all_embeddings$sim)[c(3,2,4,1)], ~{
  ggplot(.x, aes(x=V1, y=V2, col=method)) +
    geom_point() +
    facet_wrap(~method, scales = 'free', nrow = 1) +
    xlab('') + ylab('') +
    theme_clean(base_size = 14) +
    theme(plot.background = element_blank(),
          legend.background = element_blank(),
          legend.position = 'none') +
    scale_color_brewer(type = 'qual', palette = 6) +
    ggtitle(.y)
})


patchwork::wrap_plots(pp, ncol = 1, guides = 'collect') + patchwork::plot_annotation(tag_level = 'A')
ggsave("./figures/simulation_embeddings.png", width = 12, height = 10)
ggsave("./figures/simulation_embeddings.pdf", width = 12, height = 10)
```



